Datastorr: a workflow and package for delivering successive versions of ``evolving data'' directly into R

### 2018-10-10 ms.tex
Refine section on semantic versioning

Publication of a quality dataset -- a collection of measurements, stored in one or several files -- is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing datasets, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. Evidence of this trend is seen in the increasing numbers of standalone ``Data papers'' appearing in both standard domain-level journals and specialised data journals. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Force11-2014, Lowndes-2017, Perkel-2016, VanNoorden-2013, Kratz-2015,Wilkinson-2016, Yenni-2018}. In particular, in some areas, such as our own area of ecology and evolution, we are only starting to support the fact that many high-quality datasets used in research should be evolving entities \cite{Yenni-2018}.

@@

The vast majority of research projects are smaller and these currently rely on more generic data repositories for publishing of data. A common approach for publishing a dataset is to release it under a Digital Object Identifier (DOI) in a standalone data repository, such as DataDryad, Figshare, and Zenodo. While these platforms did not initially support versioning of datasets, all have subsequently added the ability to support multiple versions of a dataset, either under a single or different DOIs \cite{Nielsen-2017}. Yet, while these new features in principle allow users to access multiple versions of a dataset, the release, discovery, and access to multiple versions of a dataset is not always straightforward.

@@

To realize the full benefits of a versioned controlled dataset, users should be able to easily intuit the types of changes that have occurred among versions. Since software development has effectively already dealt with almost identical; problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.

Specifically, we suggest applying the theory of semantic versioning, developed for software distribution (\href{http://semver.org/}{semver.org}), to successive releases of an evolving dataset (Fig \ref{fig:semantic}). Semantic versioning uses a tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers. For example, version ``1.0.0''. Changes in the version number then signal the type of changes that occurred in the dataset (Fig \ref{fig:semantic}). Applying semantic versioning to data, a change from:
\begin{itemize}
  \item {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a ``correction'' (or ``patch''), for example a small error correction which is unlikely to break, or change in a substantial way, a user's analyses, although there may be minor changes in the results.
  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor enhancement'', for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset developer it might change the results of an analysis done with the prior version and so warrants closer inspection.
  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a ``major update'', for example improving the entire structure of the dataset, or adding of new columns. These changes are likely to change the results of most user's analyses or to break code that was written to work on the previous version of the dataset.
\end{itemize}
Any user of an evolving dataset where releases were tagged using semantic versioning, would immediately know the developer's view of the types of the magnitude of the changes between versions.  Note that as with software this taxonomy of changes is a language to more nuanced way to communicate from the developer to the user, rather than a set of fixed rules. Because users of data--like users of software--do unpredictable things, the developer may classify a change as minor, but still could be break some downstream analyses.

>>>>

Publication of a quality dataset -- a collection of measurements, stored in one or several files -- is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing datasets, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. Evidence of this trend is seen in the increasing numbers of standalone ``Data papers'' appearing in both standard domain-level journals and specialised data journals. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Force11-2014, Lowndes-2017, Perkel-2016, VanNoorden-2013, Kratz-2015,Wilkinson-2016, Yenni-2018}. In particular, in some areas, such as our own area of ecology and evolution, we are only starting to support the fact that some high-quality datasets may be evolving entities \cite{Yenni-2018}.

@@

The vast majority of research projects are smaller and these currently rely on more generic data repositories for publishing of data. A common approach for publishing a dataset is to release it under a Digital Object Identifier (DOI) in a standalone data repository, such as DataDryad, Figshare, and Zenodo. While these platforms did may not all initially support versioning of datasets, they now support multiple versions of a dataset, either under a single or different DOIs \cite{Nielsen-2017}. Yet, while these new features in principle allow users to access multiple versions of a dataset, the release, discovery, and access to multiple versions of a dataset is not always straightforward.

@@

To realize the full benefits of a versioned controlled dataset, users should be able to easily intuit the types of changes that have occurred among versions. Since software development has effectively already dealt with a similar problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.

Specifically, we suggest adapting the process of semantic versioning, developed for software distribution (\href{http://semver.org/}{semver.org}), to labelling of successive releases of an evolving dataset (Fig \ref{fig:semantic}). Semantic versioning uses a tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers. For example, version ``2.1.2''. Software developers increment particular numbers in a systematic way to communicate any changes to the public API for a particular piece of software.

There are two natural advantages of semantic versioning of adapting the process of semantic versioning for dataset development. The first is that enables natural ordering of releases. The second, is that it enables developers to signal the type and magnitude of change that occurred in the product between successive versions. Seeing a series of version numbers, users of an evolving dataset know the developer's view on the type and/or magnitude of change between versions.

Though the analogy to software is not perfect, datasets can also be thought of as having an Interface, which is the structure that allows for interacting with the actual data. For example, in tabular data the structure would be the column and row labels, and the data the actual cell contents. Changes to the Interface an then be labelled with semantic versioning.

Building off the guidelines for semantic versioning of software, we suggest the following guidelines for labelling a dataset with semantic versioning.
\begin{itemize}
  \item Clearly communicate the Interface to the dataset in the metadata or landing page. This includes file types, dataset structure, and element names.
  \item Use versions beginning with ``0.Y.Z'' to indicate products where the Interface is still in development.
  \item Version ``1.0.0'' defines the public Interface. 
  \item Once defined, increment version numbers to communicate any changes to the Interface.
  \item Increment the "Major" version when you make changes to the Interface that are likely incompatible with any code written to work with previous versions (i.e. causes an error). Such changes may include revising the file names, the structure of the dataset, or changing element names (e.g. column headers). 
  \item Increment the "Minor" version to communicate any changes to the Interface that are likely to be compatible with any code written to work with the previous versions (i.e. allows code to run without error). Such changes typically involve adding new data within the existing structure, so that the previous dataset version exists as a subset of the new version. For tabular data, this includes adding columns or rows. On the other hand, removing data should constitute a major version, as records previously relied on may no longer exists.
  \item Increment the "Patch" version to communicate correction of errors in the actual data, without any changes to the Interface. Such changes are unlikely to break, or change in a substantial way, analyses written with the previous version, although there may of course be changes in the results.
  \item Once a dataset version has been released, do not modify it. Further modifications are released under a new version number.
\end{itemize}

While these guidelines will help users in understanding the types of changes that have occurred between successive versions of a dataset, \emph{any} change in a dataset may alter the results of a users' analysis in non-trivial ways. Unlike developers of software, developers of a dataset cannot guarantee full backwards-compatibility, i.e. that certain results will remain unchanged in updated versions. We suggest responsibility for verifying how different versions of an evolving dataset influence their particular analysis or use thus always remains with the user, even if simply applying a so-called ``patch''. While further work -- and likely experience -- is needed to refine the process of semantic versioning for datasets to further develop understanding between data developers and data users of what different changes imply, semantic versioning still provides a more nuanced way to communicate from the developer to the user on the types of change they could expect.
